
\documentclass[addpoints]{exam}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage{booktabs}

% \printanswers

\newcommand{\event}{Statistics}
\newcommand{\tournament}{}
\def\type{0}

\setlength\fillinlinelength{\textwidth}
\CorrectChoiceEmphasis{\color{red}\bfseries}
\SolutionEmphasis{\color{red}\bfseries}

\setlength\answerclearance{2pt}
\pagestyle{headandfoot}
\runningheadrule
\runningheader{\event}{\tournament}{\ifnum\type=1 Class Set Test \else Full Name:\hspace{1cm} \fi}
\runningfootrule
\runningfooter{}{Page \thepage\ of \numpages}{}

\begin{document}
\begin{center}
    {\huge Grade 12 Foundations for College Math \\ \event
    \ifprintanswers \space Key
    \else \space \ifnum\type<2 Test \fi \ifnum\type=1 \textbf{(CLASS SET)} \fi
          \ifnum\type=2 Answer Sheet \fi
    \fi \\ \vspace{3mm}\tournament\vspace{1mm}}
\end{center}

\vspace{.25\textheight}

\begin{center}
    First Name: \ifprintanswers
    \underline{\hspace{3.5cm}}\textbf{KEY}\underline{\hspace{5.5cm}}\\\vspace{5mm}
    \else \underline{\hspace{10cm}}\\ \vspace{5mm} \fi
    Last Name: \underline{\hspace{10cm}}\\ \vspace{5mm}
\end{center}

\noindent\textbf{\underline{Directions:}}
\begin{itemize}
    \ifnum\type=1 \item \textbf{This test is a class set.} Do not write on this paper. \fi
    \item Show all work for full marks. Use the standard normal table provided where needed.
    \item The test is designed to be completed in 75 minutes.
\end{itemize}
\begin{center}
\textit{For grading use only}\\
\multirowgradetable{1}[pages]
\end{center}

\newpage
\begin{questions}

\section*{Part A --- Multiple Choice (15 marks)}
\vspace{5mm}

\question[1] The $z$-score formula is:
\begin{choices}
    \choice $z = \dfrac{\mu - x}{\sigma}$
    \correctchoice $z = \dfrac{x - \mu}{\sigma}$
    \choice $z = \dfrac{\sigma}{x - \mu}$
    \choice $z = (x - \mu)\sigma$
\end{choices}

\question[1] A data value with $z = 0$ is:
\begin{choices}
    \choice Below the mean
    \correctchoice Equal to the mean
    \choice One standard deviation above the mean
    \choice At the median
\end{choices}

\question[1] For a normal distribution with $\mu = 50$ and $\sigma = 10$, the $z$-score for $x = 65$ is:
\begin{choices}
    \choice $0.5$
    \choice $1.0$
    \correctchoice $1.5$
    \choice $2.0$
\end{choices}

\question[1] Approximately what percent of data in a normal distribution falls within 2 standard
deviations of the mean?
\begin{choices}
    \choice $68\%$
    \correctchoice $95\%$
    \choice $99.7\%$
    \choice $50\%$
\end{choices}

\question[1] A normal distribution has $\mu = 100$ and $\sigma = 15$. The range containing
the middle 68\% of data is:
\begin{choices}
    \choice $[70, 130]$
    \correctchoice $[85, 115]$
    \choice $[55, 145]$
    \choice $[90, 110]$
\end{choices}

\question[1] If $z = -1.5$, the area to the \textbf{left} of this $z$-score (from standard normal
table) is approximately:
\begin{choices}
    \choice $0.9332$
    \correctchoice $0.0668$
    \choice $0.5000$
    \choice $0.1587$
\end{choices}

\question[1] Exam scores are normally distributed with $\mu = 72$ and $\sigma = 9$. The percent
of students scoring above 81 is approximately:
\begin{choices}
    \choice $2.5\%$
    \choice $5\%$
    \correctchoice $16\%$
    \choice $34\%$
\end{choices}

\question[1] The 90th percentile of a standard normal distribution has $z \approx$:
\begin{choices}
    \choice $0.90$
    \choice $1.04$
    \correctchoice $1.28$
    \choice $1.65$
\end{choices}

\question[1] A margin of error of $\pm 3\%$ at 95\% confidence means:
\begin{choices}
    \choice The true value is exactly 3\% from the estimate
    \correctchoice We are 95\% confident the true value is within 3\% of the sample estimate
    \choice 3\% of the sample was wrong
    \choice The sample size is 3\%
\end{choices}

\question[1] To halve the margin of error, the sample size must be:
\begin{choices}
    \choice Doubled
    \choice Tripled
    \correctchoice Quadrupled
    \choice Halved
\end{choices}

\question[1] Which sampling method is \textbf{most} likely to produce a representative sample?
\begin{choices}
    \choice Surveying only volunteers
    \correctchoice Simple random sampling
    \choice Asking friends
    \choice Convenience sampling
\end{choices}

\question[1] A study finds that people who carry umbrellas get rained on more often. This means:
\begin{choices}
    \correctchoice Correlation does not imply causation
    \choice Umbrellas cause rain
    \choice The study has no value
    \choice Rain causes people to buy umbrellas
\end{choices}

\question[1] The standard deviation of a sampling distribution of means (standard error) is:
\begin{choices}
    \choice $\sigma$
    \correctchoice $\dfrac{\sigma}{\sqrt{n}}$
    \choice $\dfrac{\sigma}{n}$
    \choice $\sigma \cdot n$
\end{choices}

\question[1] A dataset has $\mu = 40$ and $\sigma = 5$. A value of 28 has $z =$:
\begin{choices}
    \choice $-1.6$
    \correctchoice $-2.4$
    \choice $2.4$
    \choice $1.6$
\end{choices}

\question[1] Response bias in a survey occurs when:
\begin{choices}
    \choice The sample is too large
    \correctchoice The wording of questions influences answers
    \choice Random sampling is used
    \choice The population is well-defined
\end{choices}


\section*{Part B --- Short Answer (33 marks)}

\question[6] The heights (cm) of 10 adults are:
\[162,\; 165,\; 168,\; 170,\; 172,\; 174,\; 176,\; 178,\; 181,\; 184\]
\begin{parts}
    \part[2] Calculate the mean $\mu$ and population standard deviation $\sigma$ (you may use
    $\sigma = 6.56$ cm).
    \part[2] Calculate the $z$-score for heights of 162 cm and 181 cm.
    \part[2] Which values fall more than 1.5 standard deviations from the mean? Identify them.
\end{parts}
\begin{solution}[9cm]
\textbf{(a)} $\mu = \dfrac{162+165+168+170+172+174+176+178+181+184}{10} = \dfrac{1730}{10} = \mathbf{173 \text{ cm}}$

$\sigma = \mathbf{6.56}$ cm (given)

\textbf{(b)} $z_{162} = \dfrac{162-173}{6.56} = \dfrac{-11}{6.56} \approx \mathbf{-1.68}$

$z_{181} = \dfrac{181-173}{6.56} = \dfrac{8}{6.56} \approx \mathbf{1.22}$

\textbf{(c)} $1.5\sigma = 1.5 \times 6.56 = 9.84$; range: $[173-9.84,\; 173+9.84] = [163.16,\; 182.84]$

Outside this range: $\mathbf{162}$ cm ($z = -1.68$) and $\mathbf{184}$ cm ($z = \frac{11}{6.56} \approx 1.68$).
\end{solution}

\question[6] The scores on a provincial math test are normally distributed with mean $\mu = 68$
and standard deviation $\sigma = 12$.
\begin{parts}
    \part[3] Using the 68-95-99.7 rule, find the percentage of students scoring:
    (i) between 56 and 80;\quad (ii) above 92;\quad (iii) below 44.
    \part[3] Using the $z$-table, find the percentage of students scoring between 60 and 85.
    ($z$-table values: $P(Z < -0.67) = 0.2514$; $P(Z < 1.42) = 0.9222$)
\end{parts}
\begin{solution}[9cm]
\textbf{(a)}
(i) $56 = 68 - 12 = \mu - \sigma$ and $80 = 68 + 12 = \mu + \sigma$. Within 1 std dev: $\mathbf{68\%}$

(ii) $92 = 68 + 2(12) = \mu + 2\sigma$. Above $\mu + 2\sigma$: $\dfrac{100-95}{2} = \mathbf{2.5\%}$

(iii) $44 = 68 - 2(12) = \mu - 2\sigma$. Below $\mu - 2\sigma$: $\mathbf{2.5\%}$

\textbf{(b)} $z_{60} = \dfrac{60-68}{12} = -0.67$;\quad $z_{85} = \dfrac{85-68}{12} = 1.42$

$P(60 < X < 85) = P(-0.67 < Z < 1.42) = 0.9222 - 0.2514 = \mathbf{0.6708 \approx 67.1\%}$
\end{solution}

\question[5] Exam scores are normally distributed with $\mu = 72$ and $\sigma = 8$.
\begin{parts}
    \part[3] Find the score at the 90th percentile. ($z_{0.90} = 1.28$)
    \part[2] What percentage of students scored below 60?
    ($P(Z < -1.5) = 0.0668$)
\end{parts}
\begin{solution}[7cm]
\textbf{(a)} $x = \mu + z\sigma = 72 + 1.28(8) = 72 + 10.24 = \mathbf{82.24}$

The 90th percentile score is approximately \textbf{82}.

\textbf{(b)} $z_{60} = \dfrac{60-72}{8} = -1.5$

$P(X < 60) = P(Z < -1.5) = \mathbf{0.0668 \approx 6.7\%}$
\end{solution}

\question[8] A survey of 400 students at a college found that 62\% support extending the library's
opening hours. The margin of error is $\pm 4.9\%$ at 95\% confidence.
\begin{parts}
    \part[2] Write the confidence interval and interpret it in context.
    \part[2] The college president claims the true support is below 55\%. Is this consistent with
    the survey results? Explain.
    \part[4] The researcher wants to reduce the margin of error to $\pm 2.45\%$ (half as large).
    How many students must be surveyed? The margin of error formula is
    $E = \dfrac{z^*}{\sqrt{n}}$ where $z^* = 1.96$ for 95\% confidence.
    Show your calculation.
\end{parts}
\begin{solution}[10cm]
\textbf{(a)} Confidence interval: $62\% \pm 4.9\% = (57.1\%,\; 66.9\%)$

We are 95\% confident that the true proportion of students who support extended hours is between
57.1\% and 66.9\%.

\textbf{(b)} The president's claim of below 55\% is \textbf{not consistent} with the survey. 55\% falls
below the lower bound of the confidence interval (57.1\%), so it is unlikely (less than 5\% chance)
that the true value is below 55\%.

\textbf{(c)} To halve the margin of error, quadruple the sample size:
\[E = \frac{z^*}{\sqrt{n}} \Rightarrow 0.0245 = \frac{1.96}{\sqrt{n}} \Rightarrow \sqrt{n} = \frac{1.96}{0.0245} = 80 \Rightarrow n = 80^2 = \mathbf{1\,600}\]

The researcher must survey \textbf{1,600 students}.

(Alternatively: $n = 4 \times 400 = 1\,600$ since halving $E$ requires quadrupling $n$.)
\end{solution}

\question[8] Read the following summary of a study and answer the questions below.

\textit{``A magazine surveyed 250 of its readers online and found that 78\% of respondents prefer
shopping in-store over online. The study concludes that the majority of Canadians prefer
in-store shopping.''}
\begin{parts}
    \part[2] Identify the sampling method used. Is this sample likely representative of all
    Canadians? Explain.
    \part[2] Identify \textbf{two} sources of bias in this study.
    \part[2] The headline reads: ``Shopping in stores causes higher customer satisfaction.''
    What statistical error does this headline make?
    \part[2] Suggest how the study design could be improved to produce more reliable results.
\end{parts}
\begin{solution}[10cm]
\textbf{(a)} This is \textbf{voluntary response sampling} (or convenience sampling — online survey of
magazine readers). It is \textbf{not representative} of all Canadians because only the magazine's
readers responded, and they likely share similar demographics, interests, and shopping preferences
that may not reflect the broader population.

\textbf{(b)} Two sources of bias:
\begin{enumerate}
\item \textbf{Voluntary response bias} — only motivated readers chose to respond, likely skewing results.
\item \textbf{Undercoverage bias} — the sample only includes magazine subscribers; people who do not
read this magazine are excluded.
(Also acceptable: question wording bias, self-selection bias.)
\end{enumerate}

\textbf{(c)} The headline confuses \textbf{correlation with causation}. The survey shows a preference for
in-store shopping; it does not establish that in-store shopping \textit{causes} higher satisfaction —
other factors may explain the preference.

\textbf{(d)} Improvements:
\begin{itemize}
\item Use simple random sampling from a broad, representative population (not just magazine readers).
\item Increase sample size significantly.
\item Use neutral, unbiased question wording.
\item Ensure all population groups (ages, locations, incomes) are represented (stratified sampling).
\end{itemize}
\end{solution}

\end{questions}
\end{document}
